An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2013-12-01 paper.tex
Finsihed editing.

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the early points in a finite-size chain to
avoid biases in the recovered posterior PDF due to the choice of the
initial location.  The points thus discarded are referred to as
``burn-in'' points.

@@

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
We choose these neighborhoods to be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  The size of a
neighborhood is inversely proportional to the local point density.
The proposed jumps are drawn from a piecewise-constant (constant on
each neighborhood) interpolation of the PDF.  There are various
techniques that could be used to construct the set of neighborhoods
associated with each point.

Ref.~\cite{Littenberg2009} decomposed the parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each sample was associated with the brick that
contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
proposal covers the entire model parameter space.  
However, the bricks in this algorithm do not adapt to the local
structure of the PDF.  One must either use small bricks to capture the
local structure of the PDF, placing many bricks in regions without
MCMC samples (which can increase memory management and access costs),
or use large bricks, missing the local structure of the PDF in
exchange for fewer empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions \cite{Voronoi1907} associated with each
MCMC sample.  The Voronoi region associated with a sample contains all
the parameter space points that are closer to that sample than any
other sample.  The Voronoi region decomposition into neighborhoods is,
in a sense, maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.
Also, the computational cost for computing the Voronoi regions
increases rapidly with dimensionality.  
% Apparently the exact computational cost is not known in arbitrary dimension.

Here we propose to use a decomposition of the parameter space into
neighborhoods based on a data structure called a kD-tree (see,
e.g.\ \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the boxes of Ref.~\cite{Littenberg2009}, and more
efficient in high-dimensional spaces than the Voronoi decomposition.

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively%
\footnote{The kD-tree datastructure defined here places box boundaries
  between the points.  An alternate definition common in the
  literature places box boundaries on the median point, but such a
  definition is inconvenient for our purposes.}: %
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N}$ space.  As an
illustration, box boundaries for a kD-tree constructed around a point
set that is normally distributed around the origin in two dimensions
are shown in Figure \ref{fig:kD-tree}.

@@

In this section, we demonstrate the efficiency of the kD-interpolated
jump proposal on a toy model-comparison problem.  We draw $N = 100$
simulated data points from a $N(0,1)$ Gaussian distribution, and then
ask whether these data are better described by a model where they are
Gaussian distributed with unknown mean $\mu$ and standard deviation
$\sigma$
%
\be
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
    \sigma^2} \right)\ ,
\ee
%
or by a model where they are Cauchy distributed with mode $\alpha$
and width $\beta$
%
\be
p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}\ .
\ee
%
We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
data set of 100 points, the relative uncertainty in determining the
parameters of the underlying distribution is approximately 10\%, so we
expect the posterior probabilities in the $(\mu,\sigma)$ and
$(\alpha,\beta)$ spaces to occupy only a few percent of the prior
volume.  The Cauchy distribution is much broader than the Gaussian (it
has no finite moments), so with uniform model priors, the posterior
probability for the Gaussian model over the Cauchy model is extremely
large:
%
\be
\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} \sim 10^9.
\ee
%
In order to ensure that the RJMCMC produces samples in the Cauchy
model at all, we impose a model prior that favors the Cauchy model by
$5 \times 10^8$ relative to the Gaussian.  The evidence ratio between
the models for our chosen data set with these priors is
%
\be \frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)}
\equiv r = 1.15,
\ee
%
yielding a theoretical maximum acceptance rate of inter-model jumps of
$(1+1/r)/2 = 0.93$.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{Figure2_acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$ for this data set).}
\end{figure}

@@

In the model selection example from Section \ref{sec:efficiency}, the
models have two-dimensional parameter spaces; in \cite{Farr2010} the
highest-dimensional model had five dimensions.  As the number of
dimensions increases the interpolation becomes more difficult for two
reasons.  First, the number of sample points required for a given
number of subdivisions in each dimension grows exponentially with
dimension, so for reasonable numbers of points a higher-dimensional kD
tree will have fewer subdivisions along each dimension.  Second, as
the dimensionality increases the volume ``at the corners'' of each kD
tree box becomes more significant; this problem is even more
pronounced when the distribution of points does not align with the
coordinate axes along which the tree subdivides.  In this section, we
discuss practical solutions to this issue in the context of a
single-model MCMC, concluding with an example of a successful
application to gravitational-wave parameter estimation.

To address the arbitrary orientation of points, the modified jump
proposal does not descend to the leaf-nodes of the tree, but instead
stops descending whenever the number of points in the box falls below
a threshold, $\Nbox$.  We then find the principal axes of the points
contained in this box and draw from a modified box which is aligned
with these axes and whose center is the mean of the points contained
within the rectangular box. The new point is chosen uniformly from
within the bounds of the modified box, subject to the restriction that
it also lie within the original, coordinate-aligned box.  Figure
\ref{fig:PCC} illustrates this two-box method.

@@

The forward jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N \left(V \cap \Vbox \right)} , \ee
%
where N is the total number of points in the kD-tree, $\Nbox$ is the
number of points in the box containing $\vtheta_{i+1}$, $V$ is the
volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
the volume of the principal-component box containing $\vtheta_{i+1}$.

The backward jump probability is computed in a similar fashion by
finding the principal-component box that would contain the current set
of parameters;.
%
\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
%
where $\Nbox'$ is the number of points in the original box containing
$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
and $\Vbox'$ is the volume of the principal-axes box containing
$\vtheta_i$.  Note, however, that the current parameters may lie in a
region of the kD box that does not intersect the principal axes box;
in this case, the backward jump probability is zero, and the algorithm
rejects the proposed step.  When used as the only jump proposal for an
RJMCMC, it is important that the kD tree proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used as one of several proposals
in the context of an MCMC, this is no longer a requirement.

It can also be advantageous, particularly in high dimensional spaces,
to accumulate more points into the kD tree as the MCMC explores more
of the parameter space.  Because this introduces a dependence on the
previous states of the MCMC, the jump proposal is no longer
Markovian.  However, it remains \emph{asymptotically} Markovian, which
ensures that the equilibrium distribution of the MCMC is correct.  To
efficiently insert points into the tree as the chain accumulates them,
the algorithm from Section \ref{sec:kDTree} must be modified:
\begin{enumerate}
\item Instead of partitioning the point set at its median, we now
  partition the bounding box at its geometric center along a
  particular dimension, which cycles as we descend the tree levels.
  Note that this allows for empty boxes if all the points cluster to
  one side of the domain.
\item When inserting a point, we descend the tree to a leaf node,
  which now can be either empty or contain one point.  If it is empty,
  we insert the point at this node.  If it contains one point, we
  subdivide the appropriate dimension, and place both points into the
  two sub-boxes.  If both land in the same sub-box, we continue
  subdividing, until each box in the sub-sub-sub... tree contains one
  or zero points.
\end{enumerate}
Insertion of a point is $\order{\log N}$, since it involves a constant
amount of work per tree level, and there are $\log N$ tree levels.
Neighborhood lookup proceeds as before, and is therefore also
$\order{\log N}$.

We have implemented this modified kD tree proposal as one of many jump
proposals in the \texttt{LALInferenceMCMC} sampler
\cite{Raymond2009,Raymond:2010,Raymond2012}.
\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
library, designed to sample the posterior on parameters of
compact-object binaries (masses, orbital orientation, distance, sky
position, etc) implied by their gravitational wave signals in
Earth-based gravitational wave detectors like LIGO.  The simplest such
signal has a nine-dimensional parameter space, and the posterior often
includes near-degeneracies and multiple modes that make convergence of
the MCMC chain very slow with traditional proposals.  Figure
\ref{fig:accratio} shows the acceptance ratio for the kD tree jump
proposal in such a simulation.  In spite of the very low acceptance
rate of the proposal, applying the kD tree proposal to one out of
twenty proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-interpolated jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in twenty proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals.}
\end{figure}

@@

The need to compare evidences for multiple models arises in a large
variety of physical and astronomical contexts.  In this paper, we
described a new technique that allows for efficient evidence
computations via a Reversible-Jump Markov chain Monte Carlo.  This
technique solves the usual problem of finding good inter-model jump
proposals in an RJMCMC by using a kD-tree to quickly and accurately
interpolate an approximate posterior PDF from a single-model MCMC run,
and then using this interpolated PDF to propose efficient inter-model
jumps.

We demonstrated the efficiency of this technique on a toy
model-comparison problem described in Section \ref{sec:efficiency}.
We also successfully applied this technique to the problem of
selecting the best model for the observed distribution of black-hole
X-ray binaries, as described in Ref.~\cite{Farr2010}.  In addition to
model comparison, the PDF interpolation described here can be useful
in multi-step MCMCs to improve the sampling efficiency by selecting
better jump proposal distributions or to test MCMC convergence.

>>>>

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the early points in a finite-size chain to
avoid biases in the recovered posterior PDF due to the choice of the
initial location.  The points thus discarded are referred to as
``burn-in'' points. \ilya{[Up to this paragraph, we mostly use "samples", not "samples"; for the rest of the paper, we mostly use "points", with occasional sliding back to "samples".  I don't have a preference, but it might be better to use uniform terminology.]}

@@

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
We choose these neighborhoods to be non-overlapping and to fill the parameter
space.  The size of a
neighborhood is inversely proportional to the local point density.
The proposed jumps are drawn from a piecewise-constant (constant on
each neighborhood) interpolation of the PDF.  
To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  
\ilya{[I moved sentences around to improve flow without actually changing the text -- hope that's OK.]}

There are various
techniques that could be used to construct the set of neighborhoods
associated with each point.
\cite{Littenberg2009} decompose the parameter space into
constant-volume ``bricks'' whose size is set by the typical size of
the peaks of the PDF.  Each sample is associated with the brick that
contains it, and the probability of proposing a jump into a particular brick is thus proportional to the
number of samples within that brick.  Additionally, an extra uniform jump proposal is added to allow for jumps into bricks that do not contain any points, so that the jump 
proposal covers the entire model parameter space.  
However, the bricks in this algorithm do not adapt to the local
structure of the PDF.  One must either use small bricks to capture the
local structure of the PDF, placing many bricks in regions without
MCMC samples (which can increase memory management and access costs),
or use large bricks, missing the local structure of the PDF in
exchange for fewer empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions \cite{Voronoi1907} associated with each
MCMC sample.  The Voronoi region associated with a sample contains all
the parameter space points that are closer to that sample than any
other sample.  The Voronoi region decomposition into neighborhoods is,
in a sense, maximally adaptive, in contrast to the approach of
\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi decomposition requires a metric on
parameter space, which may be difficult or impossible to define.
Also, the computational cost for computing the Voronoi regions
increases rapidly with dimensionality.  
% Apparently the exact computational cost is not known in arbitrary dimension.

Here we propose to use a decomposition of the parameter space into
neighborhoods based on a data structure called a kD-tree (see,
e.g.\ \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the boxes of \cite{Littenberg2009}, and more
efficient in high-dimensional spaces than the Voronoi decomposition.

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively%
\footnote{The kD-tree data structure defined here places box boundaries
  between the points.  An alternate definition common in the
  literature places box boundaries on the median point, but such a
  definition is inconvenient for our purposes.}: %
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., \cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N}$ space.  As an
illustration, box boundaries for a kD-tree constructed around a point
set that is normally distributed around the origin in two dimensions
are shown in Figure \ref{fig:kD-tree}.

@@

In this section, we demonstrate the efficiency of the kD-interpolated
jump proposal on a toy model-comparison problem.  We draw $N = 100$
simulated data points from a $N(0,1)$ Gaussian distribution, and then
ask whether these data are better described by a model where they are
Gaussian distributed with unknown mean $\mu$ and standard deviation
$\sigma$
%
\be
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
    \sigma^2} \right)\ ,
\ee
%
or by a model where they are Cauchy distributed with mode $\alpha$
and width $\beta$
%
\be
p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}\ .
\ee
%
We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
data set of 100 points, the relative uncertainty in determining the
parameters of the underlying distribution is approximately 10\%, so we
expect the posterior probabilities in the $(\mu,\sigma)$ and
$(\alpha,\beta)$ spaces to occupy only a few percent of the prior
volume.  The Cauchy distribution is much broader than the Gaussian (it
has no finite moments), so with \ilya{equal} model priors, the posterior
probability for the Gaussian model over the Cauchy model is extremely
large:
%
\be
\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} \sim 10^9.
\ee
%
In order to ensure that the RJMCMC produces samples in the Cauchy
model at all, we impose a model prior that favors the Cauchy model by
$5 \times 10^8$ relative to the Gaussian.  The evidence ratio between
the models for our chosen data set with these priors is
%
\be \frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)}
\equiv r = 1.15,
\ee
%
yielding a theoretical maximum acceptance rate of inter-model jumps of
$(1+1/r)/2 = 0.93$.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{Figure2_acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model \ilya{jump proposal} acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$ for this data set).}
\end{figure}

@@

In the model selection example from Section \ref{sec:efficiency}, the
models have two-dimensional parameter spaces; in \cite{Farr2010} the
highest-dimensional model had five dimensions.  As the number of
dimensions increases the interpolation becomes more difficult for two
reasons.  First, the number of sample points required for a given
number of subdivisions in each dimension grows exponentially with
dimension \ilya{number}, so for reasonable numbers of points a high-dimensional kD
tree will have few subdivisions along each dimension.  Second, as
the dimensionality increases the volume ``at the \ilya{edges /} corners'' of each kD
tree box becomes more significant; this problem is even more
pronounced when the distribution of points does not align with the
coordinate axes along which the tree subdivides.  In this section, we
discuss practical solutions to this issue in the context of a
single-model MCMC, concluding with an example of a successful
application to gravitational-wave parameter estimation.

\ilya{[There is at least one paragraph missing here.  What's this "single-model MCMC" and why does it require PDF interpolation?  Are you running several MCMCs in succession as increasingly good approximations to the desired posterior?  Are you using higher-temperature chains to provide a jump proposal guess for lower temperatures?   More details needed.]}

To address the arbitrary orientation of points, \ilya{[Points don't have an orientation -- they are points!  Did you mean something like "To address significant covariance between samples"?]} 
 the modified jump
proposal does not descend to the leaf-nodes of the tree, but instead
stops descending whenever the number of points in the box falls below
a threshold, $\Nbox$. \ilya{[Is $\Nbox$ the threshold, or the actual number of points in the box -- it's defined to be the latter below \Eref{modforward}?]}  We then find the principal axes of the points
contained in this box and draw from a modified box which is aligned
with these axes and whose center is the mean of the points contained
within the rectangular box. \ilya{[Is the box drawn so that all $\Nbox$ samples fall inside it?  I think so, but it should be stated explicitly.]}  The new point \ilya{[The proposed jump?  Maybe this would be a good distinction between sample and point -- use point for proposed jumps and samples for the PDF samples being interpolated; if this was done consistently, this would be more clear...]} is chosen uniformly from
within the bounds of the modified box, subject to the restriction that
it also lie within the original, coordinate-aligned box.  Figure
\ref{fig:PCC} illustrates this two-box method.

@@

The forward jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N \left(V \cap \Vbox \right)} , \ee
%
where N is the total number of points in the kD-tree, $\Nbox$ is the
number of points in the box containing $\vtheta_{i+1}$, $V$ is the
volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
the volume of the principal-component \ilya{[First time that "principal-component" is used to identify the diagonal sub-box...]} box containing $\vtheta_{i+1}$.

The backward jump probability is computed in a similar fashion by
finding the principal-component box that would contain the current set
of parameters: \ilya{[Probably not necessary -- we don't explicitly provide backward jump proposals in earlier sections]}
%
\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
%
where $\Nbox'$ is the number of points in the original box containing
$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
and $\Vbox'$ is the volume of the principal-axes box containing
$\vtheta_i$.  Note, however, that the current parameters may lie in a
region of the kD box that does not intersect the principal axes box;
in this case, the backward jump probability is zero, and the algorithm
rejects the proposed step.  When used as the only jump proposal for an
RJMCMC, it is important that the kD tree proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used as one of several proposals
in the context of an MCMC, this is no longer a requirement.

It can also be advantageous, particularly in high-dimensional spaces,
to accumulate more points into the kD tree as the MCMC explores more
of the parameter space.  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
Because this introduces a dependence on the
previous states of the MCMC, the jump proposal is no longer
Markovian.  However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference? -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance? (again, a reference?)]} , which
ensures that the equilibrium distribution of the MCMC is correct.  To
efficiently insert points into the tree as the chain accumulates them,
the algorithm from Section \ref{sec:kDTree} must be modified:
\begin{enumerate}
\item Instead of partitioning the point set at its median, we now
  partition the bounding box at its geometric center along a
  particular dimension, which cycles as we descend the tree levels.
  Note that this allows for empty boxes if all the points cluster to
  one side of the domain.
\item When inserting a point, we descend the tree to a leaf node,
  which now can be either empty or contain one point.  If it is empty,
  we insert the point at this node.  If it contains one point, we
  subdivide the appropriate dimension, and place both points into the
  two sub-boxes.  If both land in the same sub-box, we continue
  subdividing, until each box in the sub-sub-sub... tree contains one
  or zero points.
\end{enumerate}
Insertion of a point is $\order{\log N}$, since it involves a constant
amount of work per tree level, and there are $\log N$ tree levels.  \ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]}
Neighborhood lookup proceeds as before, and is therefore also
$\order{\log N}$.  \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]}

We have implemented this modified kD tree proposal as one of many jump
proposals in the \texttt{LALInferenceMCMC} sampler
\cite{vanderSluys:2008a,Raymond:2010,Raymond2012}.
\texttt{LALInferenceMCMC} is a MCMC code \cite{lalinf_method}, based on the LIGO algorithms
library, designed to sample the posterior on parameters of \ilya{merging}
compact-object binaries (masses, orbital orientation, distance, sky
position, etc) \ilya{encoded in} their gravitational wave \ilya{signatures} in
Earth-based gravitational wave detectors like LIGO.  The simplest such
signal has a nine-dimensional parameter space, and the posterior often
includes near-degeneracies and multiple modes that make convergence of
the MCMC chain very slow with traditional proposals.  Figure
\ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
proposal in such a simulation.  In spite of the very low acceptance
rate of the proposal, applying the kD-tree proposal to one out of
twenty proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.  \ilya{[Again, what exactly is reported in this figure?  Results with a jump proposal based on an interpolated PDF (of what)?  A proposal that's being updated on the fly as more samples come in?]}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-interpolated jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in twenty proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]}}
\end{figure}

@@

The need to compare evidences for multiple models arises in a large
variety of physical and astronomical contexts.  In this paper, we
described a technique that allows for efficient evidence
computations via a Reversible-Jump Markov chain Monte Carlo.  This
technique solves the usual problem of finding good inter-model jump
proposals in an RJMCMC by using a kD tree to quickly and accurately
interpolate an approximate posterior PDF from a single-model MCMC run,
and then proposing efficient inter-model
jumps \ilya{from} this interpolated PDF.

We demonstrated the efficiency of this technique on a toy
model-comparison problem described in Section \ref{sec:efficiency}.
We also successfully applied this technique to the problem of
selecting the best model for the observed distribution of black-hole
X-ray binary \ilya{masses}, as described in \cite{Farr2010}.  In addition to
model comparison, the PDF interpolation described here can be useful
in multi-step MCMCs to \sout{improve sampling efficiency by selecting
better jump proposal distributions} \ilya{propose jumps that can sample the parameter space more efficiently} or to test MCMC convergence. \ilya{[What are multi-step MCMCs (first mention)?  How do you use interpolation to test convergence?  You can't just introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}
